Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-09-09 analysis/paper/paper.Rmd
Minor edits, cleaning of R scripts

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987; and map in Figure \@ref(fig:shoredate)A]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the relative sea-level change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

In the same way is with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above. Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west/north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each radiocarbon date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites. These intervals were assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).

@@

@@

This paper has presented a new methodology for assembling and assessing the sum of many shoreline dates. This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021]. The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types.
<!-- and their potential association with overall developments in population density and mobility patterns. -->

@@

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @jorgensen2020; @ordonez2022; @lundstrom2023]. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018].

Seeing as the implemented method for shoreline dating was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application. While we believe the approach represents an improvement over previous approaches, the results should therefore be considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  By developing a methodology for handling a larger number of shoreline dates, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both proxies have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicates a largely diverging development of the two proxies through the Mesolithic. The number of shoreline dated sites undergoes some process of decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. Although this will require further substantiation, we suggest that while not devoid of a demographic signal, mobility patterns is the main determinant for the number of shoreline dated sites considered in the study. Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data is more heavily influenced by population dynamics.
keywords: |
  Mesolithic; Radiocarbon dates; Shoreline dates; Fennoscandia 
# highlights: |
#   These are the highlights. 
---

@@

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden for the Museum of Cultural History. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987, and map in Figure \@ref(fig:shoredate)A]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the relative sea-level change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

In the same way as with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above. Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west/north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each radiocarbon date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites. These intervals were assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).

@@

Finally, analogous to investigatory biases that might impact the frequency of radiocarbon dates from certain time-intervals, for example resulting from targeted sampling directed towards certain periods of interest, the distribution of sites over different elevations in the landscape might be influenced by survey practices. In evaluating two larger survey projects within the present study region, @persson2014a and @solheim2017c found that the test-pitting was generally conducted in a comparable degree across elevations, but that altitudes equating to shoreline dates towards the end of the Stone Age might be under-represented [also @solheim2018, 337]. Moving forward, similar investigations conducted in a more comprehensive manner will be important for properly evaluating the impact investigatory factors might have on the distribution of shoreline dated sites over time. This will also benefit from directly accounting for the elevation at which residential and commercial infrastructure development has been undertaken, as this generally dictates where surveys are undertaken, while also relating this to the topographic characteristics of the landscape.

@@

This paper has presented a new methodology for assembling and assessing the sum of a larger number of shoreline dates. This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021]. The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types.
<!-- and their potential association with overall developments in population density and mobility patterns. -->

@@

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @hoebe2023; @jorgensen2020; @ordonez2022; @lundstrom2023]. To account for the confounding effects these might have, a more principled and comprehensive evaluation of how investigatory practices might impact the SPDs would also be beneficial. While certainly also pertinent for the RSPD, this has generally been considered more extensively for radiocarbon data than for shoreline dated sites. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigatory effects that might impact the present SPDs inordinately [see e.g. @crema2022; @hoebe2023; @rick1987].

Seeing as the implemented method for shoreline dating was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application. While we believe the approach represents an improvement over previous implementations, the results should therefore be considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.
